A thermodynamic view of dusty protoplanetary disks

### 2017-09-12 response.txt

  While the referee is certainly correct that the single fluid analogy
  needs to made, we have attempted to explain the concept in the
  introduction. In any event, we hope the referee will agree to some
  authorial prerogative in the order of presentation.

* In section 6 on the numerical results, in particular parag. 6.5 and
  6.6, I wonder which growth rates are obtained with different radial
  mode numbers. It would be an advantage to see other values as in Fig
  8. In 6.5 kx is 30, does the instability change with smaller, larger
  modes? It is worse in 6.6, where kx is very large, 1800. Is it
  relevant to have such a high radial mode, that could barely be
  resolved in non linear numerical simulations? Finally, in a general,
  I think that a more physical summary of the results of the section 6
  should be provided. It is almost limited now to a description of the
  results.

  For section 6.5 we have updated Fig. 14 to include cases with
  different kx. All growth rates remain the same order of
  magnitude. The effect of a radial varation in rhod/rhog is also
  similar between different kx. However, surface modes only appear at
  high kx (as found for pure gas VSI, Barker & Latter, 2017)

  For section 6.6 we also updated Fig. 15 with more kx values. 
  We find to obtain instability with smaller kx, one needs a more
  rapidly-varying dust-to-gas ratio. We do not dwell 
  on this section because we have already shown, using the more general
  analyses in section 4.3 and 5.2, that disks with a radial variation in
  rhod/rhog are unstable if dust is vertically well-mixed.

### 2017-09-15 linear.tex
additional edits in response to guillaume's comments

We can neglect pressure gradients in Eq. \ref{spm} 
if $r\p_r\ln{P}\sim O(1)$ and $H_\epsilon/\Hgas\gg
\sqrt{\epsilon}\hgas/2\left(1+\epsilon\right)$. For example, if $\epsilon\simeq 0.01$ and
$\hgas\simeq 0.05$, then we require $H_\epsilon/\Hg\gg
2\times10^{-3}$. This can be met if the dust is not well settled. 
%(So the dust layer thickness $\Hd\gg 2
%\times10^{-3}\Hgas$ as well.) 
Then instability requires 
\begin{align}\label{dust_edge}
\left|\frac{\p\ln{\epsilon}}{\p r}\right| \gtrsim
  \frac{1}{\Htilde}\frac{\left(1+\epsilon\right)}{\sqrt{\epsilon}}.
%\simeq
 % \frac{1}{\sqrt{\epsilon}H_\epsilon}, 
\end{align}
%where the second equality applies to $\epsilon\ll 1$. 
That is, if the radial lengthscale of the dust-to-gas ratio 
is much less its vertical lengthscale, $L_\epsilon\ll O(\Htilde)$, 
then the system is  unstable.  
Taking $\epsilon\sim 0.01$,  we find that 
for thin dust layers with $H_\epsilon \simeq \Hd\ll \Hg$,  instability
requires $L_\epsilon\ll O(0.1\Hg)$, i.e. the dust-to-gas ratio must 
vary on an extremely short radial lengthscale. This might be achieved, for
example, at sharp edges associated with gaps opened by giant planets.

@@

In a future work we will perform a more detailed parameter survey to
compare the simplified one-fluid and full two-fluid frameworks in
calculating SI.

>>>>

We can neglect pressure gradients in Eq. \ref{spm} 
if $r\p_r\ln{P}\sim O(1)$ and $H_\epsilon/\Hgas\gg
\sqrt{\epsilon}\hgas/2\left(1+\epsilon\right)$. For example, if $\epsilon\simeq 0.01$ and
$\hgas\simeq 0.05$, then we require $H_\epsilon/\Hg\gg
2\times10^{-3}$. This can be met if the dust is not well settled {\bf (e.g. due to a small amount of turbulence).}  
%this turbulence cannot be too big because the analysis itself does not have diffusion/turbulence 
%large diffusion is expected to stabilize 
Then instability requires 
\begin{align}\label{dust_edge}
\left|\frac{\p\ln{\epsilon}}{\p r}\right| \gtrsim
  \frac{1}{\Htilde}\frac{\left(1+\epsilon\right)}{\sqrt{\epsilon}}.
%\simeq
 % \frac{1}{\sqrt{\epsilon}H_\epsilon}, 
\end{align}
%where the second equality applies to $\epsilon\ll 1$. 
That is, if the radial lengthscale of the dust-to-gas ratio 
is much less its vertical lengthscale, $L_\epsilon\ll O(\Htilde)$, 
then the system is  unstable.  
Taking $\epsilon\sim 0.01$,  we find that 
for thin dust layers with $H_\epsilon \simeq \Hd\ll \Hg$,  instability
requires $L_\epsilon\ll O(0.1\Hg)$, i.e. the dust-to-gas ratio must 
vary on an extremely short radial lengthscale. This might be achieved, for
example, at sharp edges associated with gaps opened by giant planets.

@@

In a future work we will perform a more detailed parameter survey to
compare the simplified one-fluid and full two-fluid frameworks in
calculating SI \citep[cf. comparisons using other problems, ][]{laibe14,price15}.

### 2017-09-15 model.tex
additional edits in response to guillaume's comments

In general the relative velocity $\bm{v}_\mathrm{d} -
\bm{v}_\mathrm{g}$ obeys a complicated evolutionary equation 
\citep[see, e.g.][]{youdin05a}. % obtained from taking the difference
%between evolutionary equations for $\bm{v}_\mathrm{g}$ and
%$\bm{v}_\mathrm{d}$. 
 However, for tightly-coupled dust
particles with $\tstop\OmK\ll 1 $, where $\OmK$ is the Keplerian orbital
frequency (since we are interested in protoplanetary disks), we can use the 
`terminal velocity approximation' to set %The dust-gas velocity difference 
%Then so that the dust-gas velocity
%difference is 
\begin{align}
  \bm{v}_\mathrm{d} - \bm{v}_\mathrm{g} = \frac{\nabla
    P}{\rhog}\tstop 
\end{align}
\citep{jacquet11}. This equation reflects the well-known effect of 
particle drift towards pressure maximum  \citep{weidenschilling77}.

@@

\subsection{Thermodynamics of dust-drag instabilities}\label{dust_work}
The physical interpretation of \S\ref{grow_osc} is here derived in
detail. Consider constant $c_s$ but $\tstop\neq0$ (so $\delta\mathcal{C}
\neq 0$). Eq. \ref{int_rel} indicates that $\sigma^2$ is generally complex, so
we may have growing oscillations, or overstability, due to dust-gas
friction. This is seen by taking the imaginary part of
Eq. \ref{int_rel} 
\begin{align}
%  s = \frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\int\rho\left(|\dd
%    v_r|^2 + |\dd v_z|^2\right)dV}, \label{thermal_instability}
  s = \frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\mathcal{I}^2}, \label{thermal_instability}
\end{align} 
assuming $\omega\neq0$. %This represents overstability due to dust-gas friction. 
Since drag ($\delta \mathcal{C}$) appears as a source term in our
effective energy equation, the quantity 
$\imag\left[\left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}\right]$
represent correlations between compression/expansion and  
heating/cooling.

@@

The physical interpretation of Eq. \ref{thermal_instability} is that 
work done by pressure forces in the dusty gas leads to growth in
oscillation amplitudes. To demonstrate this, we calculate the average
work done assuming periodic oscillations, and show that if the average
work done is positive, then the 
oscillation amplitude would actually grow.

@@

{\bf We consider perturbations such that 
  \begin{align}
 P \to P + \real\left(\Delta P e^{-\ii\sigma t}\right)
 \end{align}
 and similarly for $\rho$. Here, the Lagrangian 
 perturbation $\Delta$ of a variable $X$ is $\Delta X = \delta X + \bm{\xi}\cdot\nabla X$
 and $\bm{\xi}$ is the Lagrangian displacement ($\xi_{x,z} =  \ii\dd
 v_{x,z}/\sigma$). Inserting the above pressure and density fields
 into Eq. \ref{work_def}, and noting that only products of
 perturbations contribute to $\mathcal{W}$ after time-averaging, we
 find for periodic oscillations (time dependence $e^{\ii\omega t}$ and
 real $\omega$) that:  
}
\begin{align}
  \mathcal{W} = - \frac{\omega}{2} \int \imag\left(\Delta P
  \frac{\Delta \rho^*}{\rho}\right)dV,\label{work_real}
\end{align}
Eq. \ref{work_real} show that a phase difference between gas pressure and
the total density leads to work done
($\mathcal{W}\neq0$).

@@

The above discussion applies to any single fluid with a pressure and
density. In the case of interest --- dusty gas --- the work done 
is attributed to finite dust-gas drag. 
%For dusty gas, the work done during oscillations 
 %is attributed to finite dust-gas drag. 
The relative 
drift between gas and dust causes a phase difference between the two
components, and hence between pressure and total density.

>>>>

In general the relative velocity $\bm{v}_\mathrm{d} -
\bm{v}_\mathrm{g}$ obeys a complicated evolutionary equation 
\citep[see, e.g.][]{youdin05a}. % obtained from taking the difference
%between evolutionary equations for $\bm{v}_\mathrm{g}$ and
%$\bm{v}_\mathrm{d}$. 
 However, for tightly-coupled dust
particles with $\tstop\OmK\ll 1 $, where $\OmK$ is the Keplerian orbital
frequency (since we are interested in protoplanetary disks), we can use the 
`terminal velocity approximation' to set %The dust-gas velocity difference 
%Then so that the dust-gas velocity
%difference is 
\begin{align}
  \bm{v}_\mathrm{d} - \bm{v}_\mathrm{g} = \frac{\nabla
    P}{\rhog}\tstop\label{term_vel_approx} 
\end{align}
\citep{jacquet11}. This equation reflects the well-known effect of 
particle drift towards pressure maximum  \citep{weidenschilling77}.

@@

\subsection{Thermodynamics of dust-drag instabilities}\label{dust_work}
The physical interpretation of \S\ref{grow_osc} is here derived in
detail. Consider constant $c_s$ but $\tstop\neq0$ (so $\delta\mathcal{C}
\neq 0$). Eq. \ref{int_rel} indicates that $\sigma^2$ is generally complex, {\bf unless
the second integral on the right-hand-side is real. Thus } 
we may have growing oscillations, or overstability, due to dust-gas
friction. This is seen by taking the imaginary part of
Eq. \ref{int_rel} 
\begin{align}
%  s = \frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\int\rho\left(|\dd
%    v_r|^2 + |\dd v_z|^2\right)dV}, \label{thermal_instability}
  s = \frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\mathcal{I}^2}, \label{thermal_instability}
\end{align} 
assuming $\omega\neq0$. %This represents overstability due to dust-gas friction. 
Since drag ($\delta \mathcal{C}$) appears as a source term in our
effective energy equation, the quantity 
$\imag\left[\left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}\right]$
represent correlations between compression/expansion and  
heating/cooling.

@@

The physical interpretation of Eq. \ref{thermal_instability} is that 
work done by pressure forces in the dusty gas leads to growth {\bf ($s>0$) or decay ($s<0$)} 
%{\bf work done by pressure forces in the dusty gas modifies
in oscillation amplitudes. 
To demonstrate this, we calculate the average
work done assuming periodic oscillations, and show that if the average
work done is positive, then the oscillation amplitude would actually grow.

@@

{\bf Here we consider Lagragian perturbations such that 
  \begin{align}
 P \to P + \real\left(\Delta P e^{-\ii\sigma t}\right)
 \end{align}
is the pressure of a fluid element (and similarly for $\rho$). 
 The Lagrangian perturbation $\Delta$ of a variable $X$ is $\Delta X = \delta X + \bm{\xi}\cdot\nabla X$
 and $\bm{\xi}$ is the Lagrangian displacement ($\xi_{x,z} =  \ii\dd
 v_{x,z}/\sigma$). Inserting the above pressure and density fields
 into Eq. \ref{work_def}, and noting that only products of
 perturbations contribute to $\mathcal{W}$ after time-averaging, we
 find for periodic oscillations (time dependence $e^{\ii\omega t}$ and
 real $\omega$) that:  
}
\begin{align}
  \mathcal{W} = - \frac{\omega}{2} \int \imag\left(\Delta P
  \frac{\Delta \rho^*}{\rho}\right)dV,\label{work_real}
\end{align}
Eq. \ref{work_real} show that a phase difference between gas pressure and
the total density leads to work done
($\mathcal{W}\neq0$).

@@

The above discussion applies to any single fluid with a pressure and
density. In the case of interest --- dusty gas --- the work done 
is attributed to finite dust-gas drag. 
%For dusty gas, the work done during oscillations 
 %is attributed to finite dust-gas drag. 
The relative 
drift between gas and dust{\bf, which only exists if $\tstop\nabla P \neq 0$ (Eq. \ref{term_vel_approx}),} causes a phase difference between the two
components, and hence between pressure and total density.

### 2017-09-15 paper.tex
additional edits in response to guillaume's comments

Another approach is to model the dust population as a continuous, pressureless
fluid 
\citep{paardekooper06b,meheut12,laibe12,loren14,fu14b,surville16}. The 
hydrodynamic equations are evolved for two fluids: the gas and dust with density and
velocity $\rhog$, $\bm{v}_\mathrm{g}$ and $\rhod$, 
$\bm{v}_\mathrm{d}$, respectively. Source terms are introduced into
the momentum/energy equations to model dust-gas drag. This approach
can take advantage of existing numerical methods/codes for
simulating pure gas dynamics. A practical difficulty with this approach is the need to numerical stabilize the pressureless dust fluid with (e.g.) artificial diffusion.

@@

Another possibility, as annotated in Fig. \ref{pdv_cartoon}, is a dusty gas. The
relevant density here is the total density $\rho = \rhod + \rhog$, 
but the fluid pressure $P$ is due to gas only. If dust were 
perfectly-coupled to the gas, then $P$ and $\rho$ would be in phase, 
and there is no work done. However, for finite dust-gas drag, $\rhod$ and
$\rhog$ are not necessarily in phase, because dust particles can
drift relative to the gas.   
If this causes $P$ to lag 
$\rho$, then the positive work done would lead to growing
oscillations. Indeed, we show that this is true for the streaming
instability.

@@

\input{model}
\input{linear}
\input{result}
\input{summary}
\acknowledgements
We thank P. Loren-Aguilar for initial discussions that motivated this study. We also thank 
C. Baruteau, R. Dong, J. Fung, S. Inutsuka, W. Lyra, M. Pessah, and O. Umurhan 
for comments during the course of this project. This work is supported
by the Theoretical Institute for Advanced Research in Astrophysics  
(TIARA) based in Academia Sinica Institute of Astronomy and
Astrophysics (ASIAA), NASA Astrophysics Theory Program grant
NNX17AK59G, and  
the Steward Theory Fellowship at the University of Arizona.

>>>>

Another approach is to model the dust population as a continuous, pressureless
fluid 
\citep{barriere-fouchet05,paardekooper06b,ayliffe12,meheut12,laibe12,loren14,fu14b,surville16}. The 
hydrodynamic equations are evolved for two fluids: the gas and dust with density and
velocity $\rhog$, $\bm{v}_\mathrm{g}$ and $\rhod$, 
$\bm{v}_\mathrm{d}$, respectively. Source terms are introduced into
the momentum/energy equations to model dust-gas drag. This approach
can take advantage of existing numerical methods/codes for
simulating pure gas dynamics. A practical difficulty with this approach is the need to numerical stabilize the pressureless dust fluid with (e.g.) artificial diffusion.

@@

Another possibility, as annotated in Fig. \ref{pdv_cartoon}, is a dusty gas. The
relevant density here is the total density $\rho = \rhod + \rhog$, 
but the fluid pressure $P$ is due to gas only. If dust were 
perfectly-coupled to the gas, then $P$ and $\rho$ would be in phase, 
and there is no work done. However, for finite dust-gas drag, $\rhod$ and
$\rhog$ are not necessarily in phase, because dust particles can
drift relative to the gas{\bf, i.e. the dust-to-gas ratio evolves in time}. 
If this causes $P$ to lag 
$\rho$, then the positive work done would lead to growing
oscillations. Indeed, we show that this is true for the streaming
instability.

@@

\input{model}
\input{linear}
\input{result}
\input{summary}
\acknowledgements
We thank P. Loren-Aguilar for initial discussions that motivated this study. We also thank 
C. Baruteau, R. Dong, J. Fung, S. Inutsuka, G. Laibe, W. Lyra, S.-J. Paardekooper, 
M. Pessah, and O. Umurhan for comments during the course of this project. 
This work is supported by the Theoretical Institute for Advanced Research in Astrophysics   
(TIARA) based in Academia Sinica Institute of Astronomy and
Astrophysics (ASIAA), NASA Astrophysics Theory Program grant
NNX17AK59G, and  
the Steward Theory Fellowship at the University of Arizona.

### 2017-09-15 summary.tex
additional edits in response to guillaume's comments

\subsection{Implications for numerical simulations}\label{dust_sims}
The dusty/adiabatic gas equivalence means that pure
gas dynamics codes can be applied to simulate locally
isothermal/polytropic dusty gas. One need not implement a separate 
equation for the dust component: 
%In fact, no
%modification is needed  
%for strictly isothermal or polytropic gas perfectly-coupled to dust:  
 the standard energy equation substitutes for the dust continuity
equation. When $c_s$ (or $K$) is not constant and/or $\tstop\neq0$ one
should add corresponding source terms in the energy equation
(e.g. Eq. \ref{poly_energy}). 
The source term associated with dust-gas drag, $\mathcal{C}$, is 
analogous to radiative diffusion \citep{price15} or thermal
conduction, which is also common in modern hydrodynamics codes.

We have taken advantage of the dusty/adiabatic gas equivalence to 
convert the popular
\textsc{Pluto}\footnote{\url{http://plutocode.ph.unito.it/}}      
hydrodynamics code \citep{mignone07} into a dusty gas dynamics code appropriate for
simulating protoplanetary disks coupled to small dust. We will apply 
this modified code ---d\textsc{Pluto}--- to study dusty disk-planet interaction (Lin et al., 
in preparation). Preliminary results indicate the simplified one-fluid 
approach can reproduce features such as dusty rings associated with 
planet-induced gaps similar to that obtained from explicit two-fluid
simulations \citep[e.g.][]{dong17}.

@@

%implications for realistic disks
{\bf In a realistic disk, dust particles settle on a timescale $
  t_\mathrm{settle} \sim 1/\tstop\OmK^2$ \citep{takeuchi02}, compared 
  with typical VSI growth timescales, $t_\mathrm{grow}\sim 
  1/h_\mathrm{g}\OmK$. This suggest that particles with $\tstop\OmK\lesssim 
  h_\mathrm{g}$ cannot settle against the VSI. On the 
  other hand, larger particles with $\tstop\OmK\gtrsim h_g$ should
  settle to form a dusty midplane. In fact, our calculations suggest that
  settling would stabilize the disk against VSI, and allow
  further settling. The result may be a quiet, dusty midplane \citep[unless non-axisymmetric
  instabilities develop][]{chiang08} with VSI-turbulent gaseous
  atmospheres. 
}

>>>>

\subsection{Implications for numerical simulations}\label{dust_sims}
{\bf The one-fluid model with the terminal velocity approximation, Eq. \ref{masseq}---\ref{momeq}, 
has been applied to simulate dusty protoplanetary disks \citep{dipierro15,ragusa17}. These studies 
explicitly evolve the dust density.  
}

{\bf However, the dusty/adiabatic gas equivalence identified in this paper
means that one need not implement a separate equation for the dust component in
locally isothermal/polytropic gas. The standard energy equation substitutes for the dust continuity 
equation. Thus pure gas dynamic codes can be used to simulate protoplanetary dusty disks. 
} %means that pure 
%gas dynamics codes can be applied to simulate locally 
%isothermal/polytropic dusty gas. One need not implement a separate 
%equation for the dust component: 
%for strictly isothermal or polytropic gas perfectly-coupled to dust:  
% the standard energy equation substitutes for the dust continuity
%equation. 
When {\bf the sound-speed} $c_s$ (or $K$) is not constant and/or the dust-gas coupling is imperfect, $\tstop\neq0$, one
should add corresponding source terms in the energy equation
(e.g. Eq. \ref{poly_energy}). 
The source term associated with dust-gas drag, $\mathcal{C}$, is 
analogous to radiative diffusion \citep{price15} or thermal 
conduction, which is also common in modern hydrodynamics codes.

We have taken advantage of the dusty/adiabatic gas equivalence to 
convert the popular
\textsc{Pluto}\footnote{\url{http://plutocode.ph.unito.it/}}      
hydrodynamics code \citep{mignone07} into a dusty gas dynamics code appropriate for
simulating protoplanetary disks coupled to small dust. {\bf In fact, our code is unaware it is modeling dusty gas.} 
We will apply 
this modified code ---d\textsc{Pluto}--- to study dusty disk-planet interaction (Lin et al., 
in preparation). 
Preliminary results {\bf show our approach} %indicate the simplified one-fluid 
reproduce features such as dusty rings associated with 
planet-induced gaps similar to that obtained from explicit two-fluid
simulations \citep[e.g.][]{dong17}.

@@

%implications for realistic disks
{\bf In a realistic disk, dust particles settle on a timescale $
  t_\mathrm{settle} \sim 1/\tstop\OmK^2$ \citep{takeuchi02}, compared 
  with typical VSI growth timescales, $t_\mathrm{grow}\sim 
  1/h_\mathrm{g}\OmK$. This suggest that particles with $\tstop\OmK\lesssim 
  h_\mathrm{g}$ cannot settle against the VSI. On the 
  other hand, larger particles with $\tstop\OmK\gtrsim h_g$ should
  settle to form a dusty midplane. In fact, our calculations suggest that
  settling would stabilize the disk against VSI, and allow
  further settling. The result may be a quiet, dusty midplane \citep[unless non-axisymmetric
  instabilities develop, ][]{chiang08} with VSI-turbulent gaseous
  atmospheres. 
}
